Safe and Robust Formal Synthesis of Digital Controllers for Continuous Plants

### 2018-03-16 automatica/paper.tex
Addressing some of Alessandro's comments

We recall a recently introduced technique called \emph{Abstract Acceleration}~\cite{JSS14,cattaruzza2015unbounded}, 
which will be key in this work for abstraction-based synthesis.  
%\note{This clashes with the model introduced above, with inputs - motivate why? I've added the motivation above by introducing the closed  loop}
Given a model of an iterative program with dynamics
%
\begin{equation}
\vec{x}_{k+1}=\mat{A_d}\vec{x}_k \text{ where } \vec{x}_k \in \mathbb{R}^p, \mat{A_d} \in \mathbb{R}^{p \times p} \text{ and } k \in \mathbb N_0,
\end{equation}
%
\textcolor{red}{[I have modified the notations with the `d' subscript as above.]}
we want to find all possible states visited over an infinite time horizon
starting at an initial set $X_0$.  
This set is known as the \emph{reach tube} of
the model:
%
\begin{equation}
\hat{X} = \{ \vec{x}_k: k \in \mathbb N_0, \vec{x}_0 \in X_0 \text{ and } \vec{x}_{k+1}=\mat{A_d}\vec{x}_k\}.
\end{equation}
%
Using Abstract Acceleration we transform this sequential reachability problem into a one-step problem.  
We first modify the problem statement by means of acceleration into
%
\begin{equation}
\hat{X} = \{ \vec{x}_k=\mat{A_d}^k\vec{x}_0 : k \in N_0 \text{ and } \vec{x}_0 \in X_0 \}, 
\end{equation}
%
and further define new model semantics and corresponding set: 
%
\begin{equation}\label{eq:aa_reachtube}
\hat{X} \subseteq \hat{X}^\sharp = \mathcal{A}X_0, \text{ such that } \bigcup_{k \in [0\ \infty)} \mat{A_d}^k \subseteq \mathcal{A}, 
\end{equation}
%
where the newly introduced \emph{abstract reach tube} $\hat{X}^\sharp$ is an
over-approximation of the reach tube of the model \cite{JSS14}.  
In this work we extend the abstract acceleration algorithm to continuous-time models. 
% and  For this reason we must first look at the acceleration of continuous-time dynamics.

@@

\textcolor{red}{[The following three subsections take too much space, plus the results are mostly known or easily derivable with simple algebra: we need to shorten and simplify it. Dario: can you give it a crack?]}

%\note{ [add?] In this section we work with fully observed models, namely where $y = x$. } 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsubsection{Accelerated\! dynamics\! for\! continuous-time\! models\! without\! inputs}\label{sec:cont_acc_no_inputs}
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% 
Given a specific time $T$, the state $\vec{x}(t=T)$ can be expressed as a
linear function of $\vec{x}_0$ as shown below.  We assume that the system
has no inputs, so its dynamics are described by
$\dot{\vec{x}}(t)=\mat{A}\vec{x}(t)$.

\begin{lemma}
The solution to the differential equation $\dot{\vec{x}}(t)=\mat{A}\vec{x}(t)$, where $\mat{A}=\mat{S}\mat{J}\mat{S}^{-1}$, and $\mat{S}$ are the generalized eigenvectors of $\mat{A}$,
evaluated at time $T$ is
%
\begin{align}
 \vec{x}_T&=\vec{x}(t=T)=\mat{A}_{T}\vec{x} (0)\\
 \mat{A}_{T}&= \mat{S}
 \left [ \begin{array}{cccc}
 e^{T\lambda_1}  & s_1\frac{T^{1}e^{T\lambda_i}}{(2)!} & \hdots  & s_i\frac{T^{p-1}e^{T\lambda_i}}{(p-i)!} \\
0 & e^{T\lambda_i}  & s_i\frac{T^{j-i}e^{T\lambda_i}}{(j-i)!} & \vdots \\
\vdots & & \ddots & \vdots \\
0 & \cdots & 0  &e^{T\lambda_i} \\
\end{array} \right ]
 \mat{S}^{-1}
 \label{eq:continuous_tube_dyn2}\\
 &\text{where } s_i=\left\{\begin{array}{cc}1&\quad gm(\lambda_i)>1\\0&\quad gm(\lambda_i)=1\end{array}\right.,\nonumber
\end{align}
%
$\lambda_i$ are the eigenvalues of $\mat{A}$, and $gm(\lambda_i)$ is the geometric multiplicity of $\lambda_i$.
%
%$\vec{x}_T$ is the state of the system $\dot{\vec{x}}(t)=\mat{A}\vec{x}(t)$ at time t=T. 
If $\mat{J}$ is diagonal and there exists a discrete dynamics matrix for a sampling time $T_d$ such that $A_d=e^{\mat{A} T_d}$, then $\vec{x}_T=A_d^{\frac{T}{T_d}}\vec{x}_0 \text{ with } T \in \mathbb{R}^+_0$ is the state of the system at time $t=T$. 
\end{lemma}

%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsubsection{Accelerated dynamics for models with parametric inputs}\label{sec:real_discrete_param_inputs}
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

Next, we are interested in studying models with parametric inputs, which are
described by $\dot{\vec{x}}(t)=\mat{A}\vec{x}(t)+\mat{B}\vec{u}(t)$, where
$\forall t\geq 0.\,\vec{u}(t)=\vec{u}$.  For such models, a state $\vec{x}_T$
is computed next.
%
\begin{lemma}
%
The state of the model
$\dot{\vec{x}}(t)=\mat{A}\vec{x}(t)+\mat{B}\vec{u}(t)$ at time $T$ given a
parametric input $\vec{u}$ is
%
\begin{align}
\vec{x}_T&=\vec{x}(t=T)=\mat{A}_T\vec{x}_0 + \mat{B}_T\vec{u}
\label{eq:continuous_tube_param}
\end{align}
%
where $\mat{B}_T=\mat{A}^{-1}(\mat{I}-\mat{A}_T)\mat{B}$, $\forall t \leq T.\, \vec{u}(t)=\vec{u}$, and $\mat{A}_T$ as in
\eqref{eq:continuous_tube_dyn2} \footnote{When $\mat{A}$ is not invertible,
we use $\left[\begin{array}{cc}\mat{A}_T&\mat{B}_T\\
0&\mat{I}\end{array}\right]=e^{\left[\begin{array}{cc}\mat{A}&\mat{B}\\0&0\end{array}\right]}$.}. 
If $\mat{J}$ is diagonal and given a sampling time $T_d$ such that
$A_d=e^{\mat{A} T_d}$, then
$\vec{x}_T=A_d^{\frac{T}{T_d}}\vec{x}_0+(\mat{I}-\mat{A}_d^{\frac{T}{T_d}}\mat{B}_d\vec{u})
\text{ with } T \in \mathbb{R}_0^+$ (where $\mat{A}_d^{\frac{T}{T_d}} =
e^{log(\mat{A}_d) \frac{T}{T_d}} = e^{\mat{A} T}$) is the state of the
system at time $T$.
%
\end{lemma}

@@

\textcolor{red}{[This paragraph is highly unclear - Dario: can you please update it?]}

Now that we have defined the dynamics of continuous time, we can use abstract acceleration techniques to evaluate
the above model such that
\begin{equation}\label{eq:aa_cont_amatrix}
\bigcup_{T \in [0, T_s], k \in N_0} (\mat{A}_{T-kT_s}-\mat{B}_{T-kT_s}\mat{K}) (\mat{A}_d-\mat{B}_d\mat{K})^k \subseteq \mathcal{A}.
\end{equation}

Abstract acceleration works by finding support functions of the set that encompasses the matrices $\mat{A}^k$.
Support functions are scalars that measure the distance of a hyperplane tangent to the set boundary to the origin.
In the case of continuous spaces, this translates to calculating the derivatives of the envelope in order to find the direction of the support vector (normal to the hyperplane), and calculating the distance to the corresponding point where the hyperplane touches the set (at which the derivative was calculated).
The direction of the vector is determined by evaluating the support function of another point in the curve,
which is known to be smaller than that of the selected point, and the chosen hyperplanes are aligned with the eigenvectors, hence we only look at eigenvalue pairs to calculate the derivatives~\cite{cattaruzza2015unbounded}.
The corresponding derivatives are:
%
\begin{enumerate}
%
\item for real eigenvalues: we first calculate the slope of the tangent
between the two progressions: $\frac{d f(k)}{d g(k)}=\frac{f'(k)}{g'(k)}$. 
Hence $\frac{d \lambda_1^k}{d \lambda_2^k}=\frac{log(\lambda_1)
\lambda_1^k}{log(\lambda_2) \lambda_2^k}$.  From this we know that the polar
angle of the support vector is $\psi = \tan^{-1}(\frac{log(\lambda_1)
\lambda_1^k}{log(\lambda_2) \lambda_2^k})+\frac{\pi}{2}$
%
\item for conjugate eigenvalue pairs: the pair forms a logarithmic spiral. 
The tangent of a spiral with polar equation $r=a e^{b\theta}$ will have a
derivative $\frac{d r}{d \theta}=b a e^{b\theta}=b r$, hence the angle of
the vector normal to the curve is $\tan^{-1}(\frac{r}{\frac{d r}{d
\theta}})+\frac{\pi}{2}=\tan^{-1}(\frac{1}{b})+\frac{\pi}{2}$
%
\item for Jordan blocks with $gm>1$: in the case of real eigenvalues, we
follow the same procedure as in the real case, but considering that the
equation for each eigenvalue is $\frac{k^j}{j!}e^{\lambda k}$ where $j$ is
the index of the term within the Jordan block (note that when $j=0$ the term
is the same as for real eigenvalues above).  The derivatives f'(k),g'(k) are
in the form $(j+k*log(\lambda))\frac{k^{j-1}}{j!}e^{\lambda k}$.
%
\item for any combination of complex eigenvalues we use the above equations
using the envelope of their magnitude (ie $|\lambda|$).
%
\end{enumerate}

@@

We next evaluate $\hat{X}^\# \models \phi_\mathit{safety}$.  If the
verification holds we have a solution, and exit the loop.  Otherwise, we
find a counterexample iteration $k$ and corresponding initial point $x_0$
for which the property does not hold, which we use to locally refine the
abstraction.  When the abstraction cannot be further refined, we provide
them to the {\sc abstract} phase.
%
%\todo{we should have a better argument here.}
%\end{enumerate} 
\item If we reach the {\sc abstract} phase, it means that the candidate solution is not valid,
  in which case we must refine the abstraction used by the synthesizer.
\begin{enumerate}
\item Find the constraints that invalidate the property
 as a set of counterexamples for the eigenvalues, which we define as $\phi_\Lambda$. This is a constraint in the spectrum i.e., transfer function) of the closed loop dynamics. 
\item We use $\phi_\Lambda$ to
  further constrain the characteristic polynomial %\blue{[again, check the coefficients of the characteristic polynomial.]I don't find any error} 
$z^n+\sum_{i=1}^n(a_i-k_i)z^{n-i}=\prod_{i=1}^n (z-\lambda_i) : |\lambda_i|<1 \wedge \lambda_i \models \phi_{\Lambda}$. These constraints correspond to specific iterations for which the system may be unsafe.
\item Pass the refined abstraction $\phi(K)$ with the new constraints and the list of iterations $k$ to the {\sc synthesize} phase.
  %, as well as a counterexample plant coefficients ($P_a$) and repeat the loop again. 
\end{enumerate} 
\end{enumerate}

>>>>

We recall a recently introduced technique called \emph{Abstract Acceleration}~\cite{JSS14,cattaruzza2015unbounded}, 
which will be key in this work for abstraction-based synthesis.  
%\note{This clashes with the model introduced above, with inputs - motivate why? I've added the motivation above by introducing the closed  loop}
Given a model of an iterative program with dynamics
%
\begin{equation}
\vec{x}_{k+1}=\mat{A_d}\vec{x}_k \text{ where } \vec{x}_k \in \mathbb{R}^p, \mat{A_d} \in \mathbb{R}^{p \times p} \text{ and } k \in \mathbb N_0,
\end{equation}
%
\textcolor{red}{[I have modified the notations with the `d' subscript as above.]}
we want to find all possible states visited over an infinite time horizon
starting at an initial set $X_0$.  
This set is known as the \emph{reach tube} of
the model:
%
\begin{equation}
\hat{X} = \{ \vec{x}_k: k \in \mathbb N_0, \vec{x}_0 \in X_0 \text{ and } \vec{x}_{k+1}=\mat{A_d}\vec{x}_k\}.
\end{equation}
%
Using Abstract Acceleration we transform this sequential reachability problem into a one-step problem.  
We first modify the problem statement by means of acceleration into
%
\begin{equation}
\hat{X} = \{ \vec{x}_k=\mat{A_d}^k\vec{x}_0 \mid k \in N_0 \text{ and } \vec{x}_0 \in X_0 \}, 
\end{equation}
%
and further define new model semantics and corresponding set: 
%
\begin{equation}\label{eq:aa_reachtube}
\hat{X} \subseteq \hat{X}^\sharp = \mathcal{A}X_0, \text{ such that } \bigcup_{k \in [0\ \infty)} \mat{A_d}^k \subseteq \mathcal{A}, 
\end{equation}
%
where the newly introduced \emph{abstract reach tube} $\hat{X}^\sharp$ is an
over-approximation of the reach tube of the model \cite{JSS14}.  
In this work we extend the abstract acceleration algorithm to continuous-time models. 
% and  For this reason we must first look at the acceleration of continuous-time dynamics.

@@

\textcolor{red}{[The following three subsections take too much space, plus the results are mostly known or easily derivable with simple algebra: we need to shorten and simplify it. Dario: can you give it a crack?]}
\textcolor{blue}{[I merged the first two]}

%\note{ [add?] In this section we work with fully observed models, namely where $y = x$. } 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsubsection{Accelerated\! dynamics\! for\! continuous-time\! models\! with\! parametric\! inputs}\label{sec:real_discrete_param_inputs}
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% 
Given a specific time $T$, the state $\vec{x}(t=T)$ can be expressed as a
linear function of $\vec{x}_0$ and $\vec{u}$ as shown below.

\begin{lemma}
The solution to the differential equation $\dot{\vec{x}}(t)=\mat{A}\vec{x}(t)+\mat{B}\vec{u}(t)$, where
$\forall t\geq 0.\,\vec{u}(t)=\vec{u}$, $\mat{A}=\mat{S}\mat{J}\mat{S}^{-1}$, and $\mat{S}$ are the generalized eigenvectors of $\mat{A}$,
evaluated at time $T$ is
%
\begin{align}
\label{eq:continuous_tube_param}
\vec{x}_T&=\vec{x}(t=T)=\mat{A}_T\vec{x}_0 + \mat{B}_T\vec{u}\\
 \mat{A}_{T}&= \mat{S}
 \left [ \begin{array}{cccc}
 e^{T\lambda_1}  & s_1\frac{T^{1}e^{T\lambda_i}}{(2)!} & \hdots  & s_i\frac{T^{p-1}e^{T\lambda_i}}{(p-i)!} \\
0 & e^{T\lambda_i}  & s_i\frac{T^{j-i}e^{T\lambda_i}}{(j-i)!} & \vdots \\
\vdots & & \ddots & \vdots \\
0 & \cdots & 0  &e^{T\lambda_i} \\
\end{array} \right ]
 \mat{S}^{-1}
 \label{eq:continuous_tube_dyn2}\\
 \mat{B}_T&=\mat{A}^{-1}(\mat{I}-\mat{A}_T)\mat{B}
 \footnote{When $\mat{A}$ is not invertible,
we use $\left[\begin{array}{cc}\mat{A}_T&\mat{B}_T\\
0&\mat{I}\end{array}\right]=e^{\left[\begin{array}{cc}\mat{A}&\mat{B}\\0&0\end{array}\right]}$.}\\
 &\text{where } s_i=\left\{\begin{array}{cc}1&\quad gm(\lambda_i)>1\\0&\quad gm(\lambda_i)=1\end{array}\right.,\nonumber
\end{align}
%
$\lambda_i$ are the eigenvalues of $\mat{A}$, and $gm(\lambda_i)$ is the geometric multiplicity of $\lambda_i$.
%

If $\mat{J}$ is diagonal and there exists a discrete dynamics matrix for a sampling time $T_d$ such that $A_d=e^{\mat{A} T_d}$, then 
$$\vec{x}_T=A_d^{\frac{T}{T_d}}\vec{x}_0+(\mat{I}-\mat{A}_d^{\frac{T}{T_d}})\mat{B}_d\vec{u}
\text{ with } T \in \mathbb{R}_0^+$$
where $\mat{A}_d^{\frac{T}{T_d}} =e^{log(\mat{A}_d) \frac{T}{T_d}} = e^{\mat{A} T}$,
is the state of the system at time $t=T$. 
\end{lemma}

@@

\textcolor{red}{[This paragraph is highly unclear - Dario: can you please update it?]}
\textcolor{blue}{[is  this better?]}

Just as the accelerated dynamics in equation \eqref{eq:feedback_sampled_cont} can be abstacted using \eqref{eq:aa_reachtube} (where $\bigcup_{k \in N_0} (\mat{A}_d-\mat{B}_d\mat{K}) ^k \subseteq \mathcal{A}$),
we use the same principle on equation \eqref{eq:feedback_cont} to obtain 
\begin{equation}\label{eq:aa_cont_amatrix}
\bigcup_{T \in [0, T_s], k \in N_0} (\mat{A}_{T-kT_s}-\mat{B}_{T-kT_s}\mat{K}) (\mat{A}_d-\mat{B}_d\mat{K})^k \subseteq \mathcal{A}.
\end{equation}

We describe now how to obtain such a set. Abstract acceleration works by finding support functions of the set that encompasses the matrices $\mat{A}^k$.
Support functions are scalars that measure the distance of a hyperplane tangent to the set boundary to the origin.
In the case of continuous spaces, this translates to calculating the derivatives of the envelope in order to find the direction of the support vector (normal to the hyperplane), and calculating the distance to the corresponding point where the hyperplane touches the set (at which the derivative was calculated).
The direction of the vector is determined by evaluating the support function of another point in the curve,
which is known to be smaller than that of the selected point, and the chosen hyperplanes are aligned with the eigenvectors, hence we only look at eigenvalue pairs to calculate the derivatives~\cite{cattaruzza2015unbounded}.
The corresponding derivatives are:
%
\begin{enumerate}
%
\item for real eigenvalues: we first calculate the slope of the tangent
between the two progressions: $\frac{d f(k)}{d g(k)}=\frac{f'(k)}{g'(k)}$. 
Hence $\frac{d \lambda_1^k}{d \lambda_2^k}=\frac{log(\lambda_1)
\lambda_1^k}{log(\lambda_2) \lambda_2^k}$.  From this we know that the polar
angle of the support vector is $\psi = \tan^{-1}(\frac{log(\lambda_1)
\lambda_1^k}{log(\lambda_2) \lambda_2^k})+\frac{\pi}{2}$
%
\item for conjugate eigenvalue pairs: the pair forms a logarithmic spiral. 
The tangent of a spiral with polar equation $r=a e^{b\theta}$ will have a
derivative $\frac{d r}{d \theta}=b a e^{b\theta}=b r$, hence the angle of
the vector normal to the curve is $\tan^{-1}(\frac{r}{\frac{d r}{d
\theta}})+\frac{\pi}{2}=\tan^{-1}(\frac{1}{b})+\frac{\pi}{2}$
%
\item for Jordan blocks with $gm>1$: in the case of real eigenvalues, we
follow the same procedure as in the real case, but considering that the
equation for each eigenvalue is $\frac{k^j}{j!}e^{\lambda k}$ where $j$ is
the index of the term within the Jordan block (note that when $j=0$ the term
is the same as for real eigenvalues above).  The derivatives f'(k),g'(k) are
in the form $(j+k*log(\lambda))\frac{k^{j-1}}{j!}e^{\lambda k}$.
%
\item for any combination of complex eigenvalues we use the above equations
using the envelope of their magnitude (ie $|\lambda|$).
%
\end{enumerate}

@@

We next evaluate $\hat{X}^\# \models \phi_\mathit{safety}$.  If the
verification holds we have a solution, and exit the loop.  Otherwise, we
find a counterexample iteration $k$ and corresponding initial point $x_0$
for which the property does not hold, which we use to locally refine the
abstraction.  When the abstraction cannot be further refined, we provide
them to the {\sc abstract} phase.
%
%\todo{we should have a better argument here.}
%\end{enumerate} 
\item If we reach the {\sc abstract} phase, it means that the candidate solution is not valid,
  in which case we must refine the abstraction used by the synthesizer.
\begin{enumerate}
\item Find the constraints that invalidate the property
 as a set of counterexamples for the eigenvalues, which we define as $\phi_\Lambda$. This is a constraint in the spectrum i.e., transfer function) of the closed loop dynamics. 
\item We use $\phi_\Lambda$ to
  further constrain the characteristic polynomial %\blue{[again, check the coefficients of the characteristic polynomial.]I don't find any error} 
$z^n+\sum_{i=1}^n(a_i-k_i)z^{n-i}=\prod_{i=1}^n (z-\lambda_i) \text{ where } |\lambda_i|<1 \wedge \lambda_i \models \phi_{\Lambda}$. These constraints correspond to specific iterations for which the system may be unsafe.
\item Pass the refined abstraction $\phi(K)$ with the new constraints and the list of iterations $k$ to the {\sc synthesize} phase.
  %, as well as a counterexample plant coefficients ($P_a$) and repeat the loop again. 
\end{enumerate} 
\end{enumerate}
